F\# Data: \textnormal{Making structured data first-class citizens}

### 2015-02-24 icfp-2015/paper.tex
Add some type checking rules

% TODO: The model here does not actually use type erasure!!
% TODO: Maybe get rid of \mu_opt for records everywhere
% TODO: Disambigurate 's' as structural value and 's' as string
% TODO: Explain why we can only provide simple classes (!)
% TODO: Replace \nu_opt with \nu_? or something else that does not look shit

@@

\caption{Remaining reduction rules for Featherweight F\#}
\label{fig:ff-reduction}
\vspace{-0.5em}
\end{figure}

@@

The semantics fragment introduced in the previous section discusses values and operations that are used by
F\# Data. This section adds a minimal subset of the standard F\# language. We focus on the object system
(which is what F\# type providers produce) and so the following is based on Featherweight Java. However, we
only need class types (without inheritance) with properties. A class has a single implicit constructor and 
the declaration closes over constructor parameters. In addition, we include special constructs for working
with option types and lists.
%
\begin{equation*}
\begin{array}{rcl}
 \tau &\narrow{=}& \ident{int} \lsep \ident{decimal} \lsep \ident{float} \lsep \ident{bool} \lsep \ident{string} \\[0.0em]
      &\narrow{|}& C \lsep \ident{StructVal} \lsep \ident{list}\langl\tau\rangl \lsep \ident{option}\langl\tau\rangl \\[0.6em]
 L &\narrow{=}& \kvd{type}~C(\overline{x:\tau}) = \overline{M} \\[0.0em]
 M &\narrow{=}& \kvd{member}~N:\tau=e \\[0.6em]
 v &\narrow{=}& s \lsep \ident{None} \lsep \ident{Some}(v) \lsep \kvd{new}~C(\overline{v}) \\[0.0em]
 e &\narrow{=}& op \lsep e.N \lsep \kvd{new}~C(\overline{e}) \lsep {\kvd{if}~e_1~\kvd{then}~e_2~\kvd{else}~e_3}\\
   &\narrow{|}& \ident{Some}(e)\lsep\kvd{match}~e~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\
   &\narrow{|}& \kvd{match}~e~\kvd{with}~[x_1; \ldots; x_n] \rightarrow e_1 \,|\, \_ \rightarrow e_2 \\
   &\narrow{|}& [e_1;\ldots;e_n]\lsep \ident{List.map}~(\lambda x\rightarrow e_1)~e_2
\end{array}
\end{equation*}

\noindent
In the above definition, $\tau$ denotes types including a special type \ident{StructVal} for all structural 
values $s$; a class definition $L$ consists of a constructor and zero or more members. Values $v$ include 
previously defined structural values $s$ and values for the option type; finally expressions $e$ include all 
previously defined operations $op$, class construction, member access, conditionals and expressions for working 
with option values and lists. We include \ident{List.map} as a special construct to avoid making the language too complex.

In this paper, we do not define the typing of the F\# subset introduced above as our goal is not to prove
type safety for the core language. For the relativized soundness proof in Section~\ref{sec:safety}, it is
sufficient to discuss reduction.

@@

@@

Before concluding the section on formalizing type providers,

\begin{equation*}
\begin{array}{rclcrcl}
 \semalt{0} &\narrow{=}& \ident{bit} &\qquad& \semalt{1} &\narrow{=}& \ident{bit} \\
 \semalt{i} &\narrow{=}& \ident{int} && \semalt{d} &\narrow{=}& \ident{decimal} \\
 \semalt{f} &\narrow{=}& \ident{float} && \semalt{t} &\narrow{=}& \ident{string} \\
 \semalt{b} &\narrow{=}& \ident{bool} && \semalt{\kvd{null}}  &\narrow{=}& \kvd{null} \\[0em]
\end{array}
\end{equation*}
\noindent
\begin{equation*}
\begin{array}{l}
 \semalt{[s_1; \ldots; s_n]} = [\sigma_n] \quad\textnormal{where}\\[0.1em]
 \qquad\sigma_0 = \top,~ \forall i\in \{ 1.. n \}.~ \sigma_{i-1} \triangledown \semalt{s_i} \vdash \sigma_i
 \\[0.6em]
 \semalt{\nu_{\ident{opt}}~\{ \nu_1 \mapsto s_1, \ldots, \nu_n \mapsto s_n \}} =\\[0.1em]
 \qquad\nu_{\ident{opt}}~\{ \nu_1:\semalt{s_1}, \ldots, \nu_n :\semalt{s_n} \}
\end{array}
\end{equation*}

@@

\section{Relativized type safety}
\label{sec:safety}
~

We presented the system together with its syntactic equational theory and categorical semantics.
The equational theory is presented in order to explain how the systems work, but it also
provides a basis for an operational semantics for concrete systems. Exploring these, 
and their connection to the denotational semantics, is further work.

% - Safety property

@@

>>>>

% TODO: The model here does not actually use type erasure!!
% TODO: Maybe get rid of \mu_opt for records everywhere
% TODO: Disambigurate 's' as structural value and 's' as string
% TODO: Explain why we can only provide simple classes (!)
% TODO: Replace \nu_opt with \nu_? or something else that does not look shit
% TODO: Cleanup "strucutred" vs "structural"
% TODO: Call it "relativized safety" or "relativized soundness" consistently

@@

\caption{Featherweight F\# -- Remaining reduction rules}
\label{fig:ff-reduction}
\vspace{-0.5em}
\end{figure}

@@

The semantics fragment introduced in the previous section discusses values and operations that are used by
F\# Data. This section adds a minimal subset of the standard F\# language.

We focus on the object system (which is what F\# type providers produce) and so the following combines 
standard ML [?] features with Featherweight Java [?] classes. However, we only need classes with properties and
without inheritance. A class has a single implicit constructor and the declaration closes over 
constructor parameters. To avoid including all of ML, we only select constructs for working with options and lists
that we need later.
%
\begin{equation*}
\begin{array}{rcl}
 \tau &\narrow{=}& \ident{int} \lsep \ident{decimal} \lsep \ident{float} \lsep \ident{bool} \lsep \ident{string} \\[0.0em]
      &\narrow{|}& C \lsep \ident{StructVal} \lsep \ident{list}\langl\tau\rangl \lsep \ident{option}\langl\tau\rangl \\[0.6em]
 L &\narrow{=}& \kvd{type}~C(\overline{x:\tau}) = \overline{M} \\[0.0em]
 M &\narrow{=}& \kvd{member}~N:\tau=e \\[0.6em]
 v &\narrow{=}& s \lsep \ident{None} \lsep \ident{Some}(v) \lsep \kvd{new}~C(\overline{v}) \lsep [v_1; \ldots; v_n] \\[0.0em]
 e &\narrow{=}& op \lsep e.N \lsep \kvd{new}~C(\overline{e}) \lsep {\kvd{if}~e_1~\kvd{then}~e_2~\kvd{else}~e_3}\\
   &\narrow{|}& \ident{Some}(e)\lsep\kvd{match}~e~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\
   &\narrow{|}& \kvd{match}~e~\kvd{with}~[x_1; \ldots; x_n] \rightarrow e_1 \,|\, \_ \rightarrow e_2 \\
   &\narrow{|}& [e_1;\ldots;e_n]\lsep \ident{List.map}~(\lambda x\rightarrow e_1)~e_2
\end{array}
\end{equation*}

\noindent
In the above definition, $\tau$ denotes types including a type of all structured values $s$ called
\ident{StructVal}. A class definition $L$ consists of a constructor and zero or more members. Values $v$ include 
previously defined structural values $s$ and values for the option and list type; finally expressions $e$ include 
previously defined operations $op$, class construction, member access, conditionals and expressions for working 
with option values and lists. We include \ident{List.map} as a special construct to avoid making the language too complex.

Next, we define the reduction relation and (a fragment of) type checking for Featherweight F\#.
The language presented here is intentionally incomplete. We only define parts needed to prove
the relativized soundness property in Section~\ref{sec:safety}.

@@

\begin{figure}
\begin{equation*}
\qquad
\inference
  {~}
  {\Gamma \vdash s : \ident{StructVal}}
\qquad  
\inference
  {~}
  {\Gamma \vdash n : \ident{int}}
\end{equation*}

\begin{equation*}
\inference
  {\Gamma \vdash e : C & \kvd{type}~C(\overline{x:\tau}) = ..\;\kvd{member}~N_i : \tau_i = e_i\;..}
  {\Gamma \vdash e.N_i:\tau_i}
\end{equation*}

\begin{equation*}
\inference
  {\Gamma \vdash e_i : \tau_i & \kvd{type}~C(x_1:\tau_1, \ldots, x_n:\tau_n) = \ldots}
  {\Gamma \vdash \kvd{new}~C(e_1, \ldots, e_n):C}
\end{equation*}

\caption{Featherweight F\# -- Fragment of type checking}
\label{fig:ff-typecheck}
\end{figure}

\paragraph{Type checking.} 
Selected type checking rules for Featherweight F\# are shown in Figure~\ref{fig:ff-typecheck}.
The rules demonstrate the key difference between our language and standard ML or Featherweight Java:
%
\begin{itemize}
\item All structured values $s$ have a type \ident{StructVal}. Some of those have other types
  (primitive integers, decimals, floats, strings and Booleans as illustrated by the rule for $n$).
  For other values, \ident{StructVal} is the only type -- this includes records and \kvd{null}.
\item A list containing other structural values $[s_1; \ldots; s_n]$ has a type \ident{StructVal}.  
  but can also have the $\ident{list}\langl\tau\rangl$ type. Conversely, lists that contain
  non-structural values like objects or options are not of type \ident{StructVal}.
\item The two rules for checking class construction and member access are similar to corresponding
  rules of Featherweight Java.  
\end{itemize}
%
An important part of Featherweight Java that is omitted here is the checking of type declarations
(ensuring the bodies of members are well-typed). We omit this, because it is not needed for the
relativized safety theorem, which considers only classes generated by our type provider (which 
are correct by construction).

@@

Before concluding the section on formalizing type providers, there is one more missing piece. The
common supertype algorithm in Section~\ref{sec:inference} discusses the core of the type inference for
structural values, but we have not yet clarified how exactly it is used. We address this in the present
section.

Given a JSON, XML or CSV document, the F\# Data implementation constructs a structural value $s$
from the sample (this is straightforward, but Section~\ref{sec:impl-parsing} discusses some interesting 
aspects). The following defines a mapping $\semalt{-}$ which turns a sample value $s$ into a structural
type $\sigma$:
%
\begin{equation*}
\begin{array}{rclcrcl}
 \semalt{0} &\narrow{=}& \ident{bit} &\qquad& \semalt{1} &\narrow{=}& \ident{bit} \\
 \semalt{i} &\narrow{=}& \ident{int} && \semalt{d} &\narrow{=}& \ident{decimal} \\
 \semalt{f} &\narrow{=}& \ident{float} && \semalt{t} &\narrow{=}& \ident{string} \\
 \semalt{b} &\narrow{=}& \ident{bool} && \semalt{\kvd{null}}  &\narrow{=}& \kvd{null} \\[0em]
\end{array}
\end{equation*}
\noindent
\vspace{-0.2em}
\begin{equation*}
\begin{array}{l}
 \semalt{[s_1; \ldots; s_n]} = [\semalt{s_1, \ldots, s_n}]
 \\[0.5em]
 \semalt{\nu_{\ident{opt}}~\{ \nu_1 \mapsto s_1, \ldots, \nu_n \mapsto s_n \}} =\\[0.1em]
 \qquad\nu_{\ident{opt}}~\{ \nu_1:\semalt{s_1}, \ldots, \nu_n :\semalt{s_n} \}
 \\[0.5em]
 \semalt{s_1, \ldots, s_n} = \sigma_n \quad\textnormal{where}\\[0.1em]
 \qquad\sigma_0 = \top,~ \forall i\in \{ 1.. n \}.~ \sigma_{i-1} \triangledown \semalt{s_i} \vdash \sigma_i
\end{array}
\end{equation*}
%
Primitive values are mapped to their corresponding types, with the exception of $0$ and $1$, which
are inferred as \ident{bit}. For records, we return a type with field types inferred based on 
individual values.

The interesting part is inferring type based on multiple samples. We overload
the notation and write $\semalt{s_1, \ldots, s_n}$ for a type inferred from multiple samples.
This uses the common supertype relation to find a common type for all values (starting with $\top$).
This operation is used at the top-level (when calling type provider with multiple samples)
and also when inferring the type of a collection.

@@

\section{Relativized type safety}
\label{sec:safety}

Now that we have

\begin{theorem}[Relativized safety]
Given samples $s_1, \ldots, s_n$, inferred type $\sigma=\semalt{s_1, \ldots, s_n}$ and generated 
$\tau,e,L = \sem{\sigma}_x$ then:

For all $s'$ such that $\exists i.s_i :> s'$, then for any $e$ such that $x:\ident{StructVal} \vdash e:\tau$
it is the case that $e[x\leftarrow s'] \reduce^{*} v$ and $\vdash v : \tau$.

\end{theorem}
% - Safety property

@@

\label{sec:impl-parsing}
